---
title: "501_Project_2"
output:
  pdf_document: default
  html_document: default
date: "2024-12-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(lubridate)
```

```{r}
cwd <- getwd()

file_path_char <- file.path(cwd, "nc_charlotte_2020_04_01.csv")
file_path_raleigh <- file.path(cwd, "nc_raleigh_2020_04_01.csv")

charlotte <- read.csv(file_path_char)
raleigh <- read.csv(file_path_raleigh)
```


```{r}
summary(charlotte)
summary(raleigh)

str(charlotte)
str(raleigh)

nrow(charlotte)
nrow(raleigh)
```
```{r}
head(charlotte$date, 10)
head(raleigh$date, 10)

range(charlotte$date, na.rm = TRUE)
range(raleigh$date, na.rm = TRUE)

sum(is.na(charlotte$date))
sum(is.na(raleigh$date))
```
```{r}
charlotte$date <- as.Date(charlotte$date, format = "%Y-%m-%d")
raleigh$date <- as.Date(raleigh$date, format = "%Y-%m-%d")
str(charlotte$date)
str(raleigh$date)

```



```{r}
charlotte$age_group <- cut(
  charlotte$subject_age,
  breaks = c(-Inf, 25, 50, Inf),
  labels = c("16-25", "25-50", "50+"),
  right = FALSE
)
raleigh$age_group <- cut(
  raleigh$subject_age,
  breaks = c(-Inf, 25, 50, Inf),
  labels = c("16-25", "25-50", "50+"),
  right = FALSE
)

```

```{r}
charlotte_clean <- charlotte %>%
  filter(!is.na(subject_sex) & !is.na(age_group))
raleigh_clean <- raleigh %>%
  filter(!is.na(subject_sex) & !is.na(age_group))
```

```{r}
nrow(charlotte_clean)
nrow(raleigh_clean)

```
```{r}
charlotte_demographics <- table(charlotte_clean$subject_sex, charlotte_clean$age_group)

raleigh_demographics <- table(raleigh_clean$subject_sex, raleigh_clean$age_group)

charlotte_demographics
raleigh_demographics

```
```{r}
charlotte_outcomes <- charlotte_clean %>%
  group_by(subject_sex, age_group) %>%
  summarise(
    arrests = sum(arrest_made, na.rm = TRUE),
    citations = sum(citation_issued, na.rm = TRUE),
    warnings = sum(warning_issued, na.rm = TRUE),
    total_stops = n()
  )

raleigh_outcomes <- raleigh_clean %>%
  group_by(subject_sex, age_group) %>%
  summarise(
    arrests = sum(arrest_made, na.rm = TRUE),
    citations = sum(citation_issued, na.rm = TRUE),
    warnings = sum(warning_issued, na.rm = TRUE),
    total_stops = n()
  )

charlotte_outcomes
raleigh_outcomes

```

```{r}
charlotte_proportions <- charlotte_clean %>%
  group_by(subject_sex, age_group) %>%
  summarise(
    arrest_rate = sum(arrest_made, na.rm = TRUE) / n(),
    citation_rate = sum(citation_issued, na.rm = TRUE) / n(),
    warning_rate = sum(warning_issued, na.rm = TRUE) / n(),
    .groups = "drop"
  )

raleigh_proportions <- raleigh_clean %>%
  group_by(subject_sex, age_group) %>%
  summarise(
    arrest_rate = sum(arrest_made, na.rm = TRUE) / n(),
    citation_rate = sum(citation_issued, na.rm = TRUE) / n(),
    warning_rate = sum(warning_issued, na.rm = TRUE) / n(),
    .groups = "drop"
  )

charlotte_proportions
raleigh_proportions
```
Gender Disparities:
1. Males have higher arrest rates across all age groups, indicating possible policing disparities.
2. Females generally receive more warnings and slightly fewer citations compared to males.

Age-Related Trends:
1. Arrest rates decrease with age for both genders.
2. Citation rates remain stable across age groups but show slight declines with increasing age.
3. Warnings increase with age for both genders, especially for females.

```{r}

charlotte_proportions$city <- "Charlotte"
raleigh_proportions$city <- "Raleigh"
combined_proportions <- rbind(charlotte_proportions, raleigh_proportions)

ggplot(combined_proportions, aes(x = age_group, fill = subject_sex)) +
  geom_bar(aes(y = arrest_rate), stat = "identity", position = "dodge") +
  facet_wrap(~ city) +
  labs(
    title = "Arrest Rate by Gender and Age Group",
    x = "Age Group",
    y = "Arrest Rate",
    fill = "Gender"
  ) +
  theme_minimal()

ggplot(combined_proportions, aes(x = age_group, fill = subject_sex)) +
  geom_bar(aes(y = citation_rate), stat = "identity", position = "dodge") +
  facet_wrap(~ city) +
  labs(
    title = "Citation Rate by Gender and Age Group",
    x = "Age Group",
    y = "Citation Rate",
    fill = "Gender"
  ) +
  theme_minimal()

ggplot(combined_proportions, aes(x = age_group, fill = subject_sex)) +
  geom_bar(aes(y = warning_rate), stat = "identity", position = "dodge") +
  facet_wrap(~ city) +
  labs(
    title = "Warning Rate by Gender and Age Group",
    x = "Age Group",
    y = "Warning Rate",
    fill = "Gender"
  ) +
  theme_minimal()

```
Hypotheses for the Chi-Squared Test:

Null Hypothesis (H$_0$): Stop outcomes are independent of gender and age group (no disparities).
Alternative Hypothesis (H$_A$): Stop outcomes are dependent on gender and age group (disparities exist).

```{r}
charlotte_table <- charlotte_clean %>%
  group_by(subject_sex, age_group) %>%
  summarise(
    arrests = sum(arrest_made, na.rm = TRUE),
    citations = sum(citation_issued, na.rm = TRUE),
    warnings = sum(warning_issued, na.rm = TRUE),
    .groups = "drop"
  )

raleigh_table <- raleigh_clean %>%
  group_by(subject_sex, age_group) %>%
  summarise(
    arrests = sum(arrest_made, na.rm = TRUE),
    citations = sum(citation_issued, na.rm = TRUE),
    warnings = sum(warning_issued, na.rm = TRUE),
    .groups = "drop"
  )

charlotte_matrix <- as.matrix(charlotte_table[, 3:5])
raleigh_matrix <- as.matrix(raleigh_table[, 3:5])

charlotte_test <- chisq.test(charlotte_matrix)
raleigh_test <- chisq.test(raleigh_matrix)

charlotte_test
raleigh_test

```
1. For both Charlotte and Raleigh, the p-value is far below the conventional significance level of 0.05.
2. This means we reject the null hypothesis: stop outcomes (arrests, citations, warnings) are not independent of gender and age group.
3. These results indicate that gender and age group play a significant role in determining stop outcomes.
4. For example, males in younger age groups may be disproportionately more likely to be arrested, as observed in the previous analyses.

```{r}
charlotte_clean$party_in_power <- ifelse(
  charlotte_clean$date < as.Date("2001-01-01"), "Democratic",
  ifelse(charlotte_clean$date >= as.Date("2001-01-01") & charlotte_clean$date <= as.Date("2008-12-31"), "Republican", "Democratic")
)

raleigh_clean$party_in_power <- ifelse(
  raleigh_clean$date < as.Date("2001-01-01"), "Democratic",
  ifelse(raleigh_clean$date >= as.Date("2001-01-01") & raleigh_clean$date <= as.Date("2008-12-31"), "Republican", "Democratic")
)

table(charlotte_clean$party_in_power)
table(raleigh_clean$party_in_power)

```
```{r}
charlotte_party_analysis <- charlotte_clean %>%
  group_by(subject_sex, age_group, party_in_power) %>%
  summarise(
    arrest_rate = sum(arrest_made, na.rm = TRUE) / n(),
    citation_rate = sum(citation_issued, na.rm = TRUE) / n(),
    warning_rate = sum(warning_issued, na.rm = TRUE) / n(),
    .groups = "drop"
  )

raleigh_party_analysis <- raleigh_clean %>%
  group_by(subject_sex, age_group, party_in_power) %>%
  summarise(
    arrest_rate = sum(arrest_made, na.rm = TRUE) / n(),
    citation_rate = sum(citation_issued, na.rm = TRUE) / n(),
    warning_rate = sum(warning_issued, na.rm = TRUE) / n(),
    .groups = "drop"
  )

charlotte_party_analysis
raleigh_party_analysis

```

```{r}
charlotte_party_table <- charlotte_clean %>%
  group_by(party_in_power) %>%
  summarise(
    arrests = sum(arrest_made, na.rm = TRUE),
    citations = sum(citation_issued, na.rm = TRUE),
    warnings = sum(warning_issued, na.rm = TRUE),
    .groups = "drop"
  )

charlotte_party_matrix <- as.matrix(charlotte_party_table[, 2:4])

charlotte_party_test <- chisq.test(charlotte_party_matrix)

raleigh_party_table <- raleigh_clean %>%
  group_by(party_in_power) %>%
  summarise(
    arrests = sum(arrest_made, na.rm = TRUE),
    citations = sum(citation_issued, na.rm = TRUE),
    warnings = sum(warning_issued, na.rm = TRUE),
    .groups = "drop"
  )

raleigh_party_matrix <- as.matrix(raleigh_party_table[, 2:4])
raleigh_party_test <- chisq.test(raleigh_party_matrix)

charlotte_party_test
raleigh_party_test

```
Null Hypothesis (H$_0$):
Stop outcomes (arrests, citations, warnings) are independent of the party in power.
Alternative Hypothesis (H$_A$):
Stop outcomes (arrests, citations, warnings) are dependent on the party in power.

The p-values for both Charlotte and Raleigh are extremely small (< 2.2e-16), leading us to reject the null hypothesis. This means:

Stop outcomes are not independent of the party in power.
```{r}
charlotte_party_analysis$city <- "Charlotte"
raleigh_party_analysis$city <- "Raleigh"
combined_party_analysis <- rbind(charlotte_party_analysis, raleigh_party_analysis)

# Arrest Rate Visualization
ggplot(combined_party_analysis, aes(x = age_group, y = arrest_rate, fill = party_in_power)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ city + subject_sex) +
  labs(
    title = "Arrest Rate by Political Party, Gender, and Age Group",
    x = "Age Group",
    y = "Arrest Rate",
    fill = "Party in Power"
  ) +
  theme_minimal()

ggplot(combined_party_analysis, aes(x = age_group, y = citation_rate, fill = party_in_power)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ city + subject_sex) +
  labs(
    title = "Citation Rate by Political Party, Gender, and Age Group",
    x = "Age Group",
    y = "Citation Rate",
    fill = "Party in Power"
  ) +
  theme_minimal()

ggplot(combined_party_analysis, aes(x = age_group, y = warning_rate, fill = party_in_power)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ city + subject_sex) +
  labs(
    title = "Warning Rate by Political Party, Gender, and Age Group",
    x = "Age Group",
    y = "Warning Rate",
    fill = "Party in Power"
  ) +
  theme_minimal()

```

