---
title: "project"
output: pdf_document
date: "2024-12-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("tidyverse")) install.packages("tidyverse")
if (!require("lubridate")) install.packages("lubridate")
if (!require("broom")) install.packages("broom")


library(tidyverse)
library(lubridate)
library(broom)

```

```{r}
charlotte_data <- suppressWarnings(read_csv("nc_charlotte_2020_04_01.csv", show_col_types = FALSE))
raleigh_data <- suppressWarnings(read_csv("nc_raleigh_2020_04_01.csv", show_col_types = FALSE))


charlotte_data <- charlotte_data %>%
  mutate(city = "Charlotte")

raleigh_data <- raleigh_data %>%
  mutate(city = "Raleigh")

combined_data <- bind_rows(charlotte_data, raleigh_data)

glimpse(combined_data)
```

```{r, echo=FALSE}
combined_data <- combined_data %>%
  mutate(
    date = ymd(date), 
    year = year(date), 
    party_in_power = case_when(
      year %in% 2000:2000 ~ "Democratic",
      year %in% 2001:2008 ~ "Republican",  
      year %in% 2009:2015 ~ "Democratic",  
      TRUE ~ NA_character_
    ),
    search_conducted = as.logical(search_conducted), 
    contraband_found = as.logical(contraband_found),
    subject_race = as.factor(subject_race)
  ) %>%
  filter(!is.na(subject_race))  


glimpse(combined_data)
sink("output.txt")
glimpse(combined_data)
sink()


```

```{r}
# Calculate search and contraband rates
search_rates <- combined_data %>%
  group_by(city, subject_race, search_conducted) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(percentage = count / sum(count) * 100)

contraband_rates <- combined_data %>%
  filter(search_conducted == TRUE) %>%
  group_by(city, subject_race, contraband_found) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(percentage = count / sum(count) * 100)

print(search_rates)
print(contraband_rates)

```

```{r}
# Search rates by political affiliation
search_rates_party <- combined_data %>%
  group_by(city, party_in_power, subject_race, search_conducted) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(percentage = count / sum(count) * 100)

# Contraband discovery rates by political affiliation
contraband_rates_party <- combined_data %>%
  filter(search_conducted == TRUE) %>%
  group_by(city, party_in_power, subject_race, contraband_found) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(percentage = count / sum(count) * 100)

print(search_rates_party)
print(contraband_rates_party)

```

Let's do a chi-squared test to see if there's a statistically significant association between race and search rates.

```{r}
# Perform the test separately for each city
search_test <- combined_data %>%
  group_by(city, subject_race, search_conducted) %>%
  summarize(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = search_conducted, values_from = count, values_fill = 0)

search_test_split <- split(search_test, search_test$city)

results <- lapply(search_test_split, function(city_data) {
  city_data <- city_data %>%
    select(-city) %>%
    column_to_rownames("subject_race")
  
  chisq.test(city_data)
})

results


```

Clearly, we notice that there is a significant difference between search rates based on the race of the person involved. Let's do a similar chi squared test and confirm this for whether contraband was found or not.

```{r}
contraband_test <- combined_data %>%
  filter(search_conducted == TRUE) %>%
  group_by(city, subject_race, contraband_found) %>%
  summarize(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = contraband_found, values_from = count, values_fill = 0)

contraband_test_split <- split(contraband_test, contraband_test$city)

contraband_results <- lapply(contraband_test_split, function(city_data) {
  city_data <- city_data %>%
    select(-city) %>%
    column_to_rownames("subject_race")
  
  suppressWarnings(chisq.test(city_data))
})

contraband_results

```

We notice that there is a statistical significance in contraband discovery rates too across both cities. This can be visualized and easily digested as seen below.

```{r}
combined_data %>%
  group_by(city, subject_race, search_conducted) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(percentage = count / sum(count) * 100) %>%
  ggplot(aes(x = subject_race, y = percentage, fill = search_conducted)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ city) +
  labs(title = "Search Rates by City and Race", y = "Percentage", x = "Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

combined_data %>%
  filter(search_conducted == TRUE) %>%
  group_by(city, subject_race, contraband_found) %>%
  summarize(count = n(), .groups = "drop") %>%
  mutate(percentage = count / sum(count) * 100) %>%
  ggplot(aes(x = subject_race, y = percentage, fill = contraband_found)) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~ city) +
  labs(title = "Contraband Discovery Rates by City and Race", y = "Percentage", x = "Race") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))


```

We now take a look at whether it matters what party was in power. As can be seen below, regardless of what party was in power, there's a significant difference in search rates between the races. This is also confirmed for contraband discovery rates.

```{r}
# Group data by race, search conducted, and party in power
search_test_party <- combined_data %>%
  group_by(party_in_power, subject_race, search_conducted) %>%
  summarize(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = search_conducted, values_from = count, values_fill = 0)

search_test_party_split <- split(search_test_party, search_test_party$party_in_power)

search_results_party <- lapply(search_test_party_split, function(party_data) {
  party_data <- party_data %>%
    select(-party_in_power) %>%
    column_to_rownames("subject_race")
  chisq.test(party_data)
})

print(search_results_party)

```

```{r}
contraband_test_party <- combined_data %>%
  filter(search_conducted == TRUE) %>%
  group_by(party_in_power, subject_race, contraband_found) %>%
  summarize(count = n(), .groups = "drop") %>%
  pivot_wider(names_from = contraband_found, values_from = count, values_fill = 0)

contraband_test_party_split <- split(contraband_test_party, contraband_test_party$party_in_power)

contraband_results_party <- lapply(contraband_test_party_split, function(party_data) {
  party_data <- party_data %>%
    select(-party_in_power) %>%
    column_to_rownames("subject_race")
  chisq.test(party_data)
})

print(contraband_results_party)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
